### 2020-09-17 paper/implementation.tex
Fixes

A repository containing the discussed source code and tests is available under the link: \href{https://github.com/AdrianSiwiec/Perfect-Graphs}{\texttt{https://github.com/AdrianSiwiec/Perfect-Graphs}}
\todo{make repo public and add readme on howto run, some script creating and running default tests}

\section{Berge graphs recognition: na\"ive approach}
When implementing an algorithm with the running time of $O(|V|^9)$ (especially one as complex as CCLSV), one of the first questions that pop up is whether this algorithm is at all usable. We couldn't find any existing programs that test whether the graph is perfect, be it an implementation of CCLSV, or a na\"ive approach.\todo{well.. there's this java impl} So, to test the usability of CCLSV implementation, we measured it against our own na\"ive perfect recognition algorithm. Let us describe it briefly.

@@

With those requirements defined a simplest algorithm would be to implement a sort of a "counter" with base of $|V(G)|$. In short: we take a path on a input, increment its last vertex until it is a neighbor of the vertex one before last. Then we check if generated sequence is a path. If it is we return it and if we run out of vertices before a path is found, we increment one before last vertex until it is a neighbor of one vertex before it, set last vertex to first and continue the process. If still no path is found, we increment vertices closer to beginning of the path, until all a path is found or we check all candidates.

@@

We have an array $first$ in which for each vertex is written its first neighbor on its neighbors list, and an array $next$ of size $|V(G)|^2$ where for each pair of vertices $a, b$ if $a$ and $b$ are connected, there is written a neighbor of $a$ that is next after $b$ in a neighbors list of $a$ (or a flag indicating $b$ is $a$'s last neighbor). Then, say our input is a path $v = v_1, \ldots v_k$. If $next[v_{k-1}][v_k]$ exists we change $v_k$ to $next[v_{k-1}][v_k]$ and return $v$. If it indicated that $v_k$ is $v_{k-1}$'s last neighbor, we set $v_{k-1}$ to $next[v_{k-2}][v_{k-1}]$ and then $v_k$ to $first[v_{k-1}]$ (or we go further back, if all neighbors of $v_{k-2}$ are done).

This simple change in data structure design gave us a speedup of overall running time of our na\"ive algorithm in the range of about 5x.

@@

A theoretically optimal solution in general would be to have a hashing set of vertices. For each vertex in a path candidate, we check if it is already in a set. If it is, return that not all vertices are unique, else add it to the set.

In our use case, a few optimizations can be made. First, we don't need a hashing set. We can have an array of bools, of size $|V(G)|$ and mark vertices of a path there. Paths are usually much shorter than $|V(G)|$, but we operate on small graphs, so we can afford this. Second, we notice that we don't need to create this array for each call of \texttt{areUnique} procedure. Let's instead have a static array $stamp$ of ints and a static $counter$ of how many times we called \texttt{areUnique} method. For each element of a path $v_i$, if the value of $stamp$ array is equal to $counter$ we report that vertices are not unique. Else we set $stamp[v_i] = counter$. When returning from \texttt{areUnique} method we increment $counter$.

@@

Because of a good performance of our optimized algorithm for generating paths, we use it whenever possible. It can be easily modified to enumerate all ''paths'' that can have additional edges or to enumerate holes and therefore has much use in the CCLSV algorithm. For example, when searching for jewels we generate with it all ''paths'' with possible additional edges, of length 5 and for each check if it has required properties of a jewel. This proved to be much faster than generating vertices and checking their properties one by one. We use the same algorithm for generating starting vertices for a possible $\T_2$ and $\T_3$. Before optimizing path enumeration algorithm, all calls of $areUnique$ alone took abour 70\% of total running time of the CCLSV algorithm. After optimizing $areUnique$, path enumeration still took more than 50\% of total running time. After all optimizations it is almost unnoticable.

@@

Callgrind is a profiling tool that records the call history and event counts (data reads, cache misses etc.) of a program and presents it as a call-graph. It is then possible to visualise this, we used a tool \emph{gprof2dot} \cite{gprof2dot} to generate visual call graphs from callgrind's output.

Before any optimizatios, a major part of Berge recognition algorithm was being spent on enumerating all possible paths of given length -- this is done either to find a simple structure or check a possible near cleaner for an amenable odd hole. Therefore optimizations of the path enumerating algorithm were also helpful in speeding up CCLSV implementation.

@@

Another algorithm for which we found major speedups is the algorithm to generate all possible near-cleaners (\Cref{alg:listNearCleaners}). We optimized it by using a a \texttt{dynamic\_bitset} from the boost library \cite{boost}. It is a data structure to represent a set of bits, that also allows for fast bitwise operators that one can apply to builtin integers. At the very end of \Cref{alg:listNearCleaners} a set $\mathcal{R}$ is constructed, which is an union of all possible unions of pairs $X_i$ and $N_j$. We used \texttt{dynamic\_bitset} to represent each of the sets $X_i$ and $N_j$, and the elements of the set $\mathcal{R}$. This significantly speeds up the operation of union of two sets (line \ref{line:listNCcalcR}). With this modification, the speedup of \Cref{alg:listNearCleaners} was about 20\%.

After these optimizations, checking each potential near-cleaner by \Cref{alg:testNearCleaner} is by far the biggest bottleneck of the CCLSV algorithm. We didn't find any major optimizations there, but the \Cref{alg:testNearCleaner} has a good potential for parrallelization, which we explore in \Cref{sec:CUDA}.

@@

CUDA is a parrallel computing platform model created by Nvidia and executed on Nvidia GPUs. It allows to run a program  by many CUDA threads at once. Although usually described by the the PRAM models, implementing PRAM algorithm in CUDA architecture is not as straightforward as implementing a RAM algorith on a CPU. Some of the causes are: synchronization between threads is very costly, there are multiple memory models to choose from, threads run divided in blocks and it is possible to efficiently share memory within blocks and copying memory from and back to CPU suffers from a large latency. What is more, there are no standard dynamic structures in the PRAM model, let alone CUDA, that are taken for granted in the RAM model, such as hashing maps or sets. 
% We used a library called \emph{moderngpu} for our implementation, because it allows a slightly cleaner code.

@@

We posed a question if it would be profitable to use GPU for the problem of recognizing Berge graphs. On one hand the graphs we are working on are small and our time complexity so large that a speedup from massively parallel architecture could be profitable. On the other hand, the algorithm is complex and not easy to parallelise. Therefore we decided to analyze the  CCLSV algorithm and parrallelize parts that would benefit from it the most.

@@

Let us recall the \Cref{alg:testNearCleaner}. It calculates array $R$ of shortest paths, then for each 3-vertex path and an additional vertex it does $O(1)$ checks to see if they along with two paths from $R$ give us an odd hole. We will parallelise the work after calculating $R$, that is, lines \ref{line:cudaStart}-\ref{line:cudaEnd}. Let us notice, that for all $X$s, all 3-vertex paths enumerated in line \ref{line:3vertex} are the same. We calculate them beforehand, then each CUDA thread receives one 3-vertex path $x_1 - x_2 - x_3$ and an additional vertex $y_1$ and performs the required checks in lines \ref{line:cudaWorkStard}-\ref{line:cudaWorkEnd}. This is almost perfect scenario for the GPU -- we do a simple SIMD work in parallel, without a lot scattered memory access.

The part of the CCLSV implemented on the GPU achieved speedup of up to 15x. Which gave us a speedup of up to 9x for overall CCLSV algorithm, depending on a class of graph we benchmark on.

>>>>

A repository containing the discussed source code and tests is available under the link: \href{https://github.com/AdrianSiwiec/Perfect-Graphs}{\texttt{https://github.com/AdrianSiwiec/Perfect-Graphs}}

\section{Berge graphs recognition: na\"ive approach}
When implementing an algorithm with the running time of $O(|V|^9)$ (especially one as complex as CCLSV), one of the first questions that pop up is whether this algorithm is at all usable. We couldn't find any existing programs that test whether the graph is perfect, be it an implementation of CCLSV, or a na\"ive approach. So, to test the usability of CCLSV implementation, we measured it against our own na\"ive perfect recognition algorithm. Let us describe it briefly.

@@

With those requirements defined a simplest algorithm would be to implement a sort of a "counter" with base of $|V(G)|$. In short: we take a path on a input, increment its last vertex until it is a neighbor of the vertex one before last. Then we check if generated sequence is a path -- all of its vertices are unique and there are edges only consecutive vertices share an edge. If it is we return it and if we run out of vertices before a path is found, we increment one before last vertex until it is a neighbor of one vertex before it, set last vertex to first and continue the process. If still no path is found, we increment vertices closer to beginning of the path, until all a path is found or we check all candidates.

@@

We have an array $first$ in which for each vertex is written its first neighbor on its neighbors list, and an array $next$ of size $|V(G)|^2$ where for each pair of connected vertices $a, b$, there is written a neighbor of $a$ that is next after $b$ in a neighbors list of $a$ (or a flag indicating $b$ is $a$'s last neighbor). Then, say our input is a path $v = v_1, \ldots v_k$. If $next[v_{k-1}][v_k]$ exists we change $v_k$ to $next[v_{k-1}][v_k]$ and return $v$. If it indicated that $v_k$ is $v_{k-1}$'s last neighbor, we set $v_{k-1}$ to $next[v_{k-2}][v_{k-1}]$ and then $v_k$ to $first[v_{k-1}]$ (or we go further back, if all neighbors of $v_{k-2}$ are done).

This simple change in the data structure design gave us a speedup of overall running time of our na\"ive algorithm in the range of about 5x.

@@

A theoretically optimal solution in general would be to have a hashing set of vertices. For each vertex in a path candidate, we check if it is already in the set. If it is, return that not all vertices are unique, else add it to the set.

In our use case, a few optimizations can be made. First, we don't need a hashing set. We can have an array of bools, of size $|V(G)|$ and mark vertices of the path there. Paths are usually much shorter than $|V(G)|$, but we operate on small graphs, so we can afford this. Second, we notice that we don't need to create this array for each call of \texttt{areUnique} procedure. Let's instead have a static array $stamp$ of ints and a static $counter$ of how many times we called \texttt{areUnique} method. For each element of a path $v_i$, if the value of $stamp$ array is equal to $counter$ we report that vertices are not unique. Else we set $stamp[v_i] = counter$. When returning from \texttt{areUnique} method we increment $counter$.

@@

Because of a good performance of our optimized algorithm for generating paths, we use it whenever possible. It can be easily modified to enumerate all ''paths'' that can have additional edges or to enumerate holes and therefore has much use in the CCLSV algorithm. For example, when searching for jewels we generate with it all ''paths'' with possible additional edges, of length 5 and for each check if it has required properties of a jewel. This proved to be much faster than generating vertices and checking their properties one by one. We use the same algorithm for generating starting vertices for a possible $\T_2$ and $\T_3$. Before optimizing path enumeration algorithm, all calls of $areUnique$ alone took about 70\% of total running time of the CCLSV algorithm. After optimizing $areUnique$, path enumeration still took more than 50\% of total running time. After all optimizations it is almost unnoticeable.

@@

Callgrind is a profiling tool that records the call history and event counts (data reads, cache misses etc.) of a program and presents it as a call-graph. It is then possible to visualize this, we used a tool \emph{gprof2dot} \cite{gprof2dot} to generate visual call graphs from callgrind's output.

Before any optimizations, a major part of Berge recognition algorithm was being spent on enumerating all possible paths of given length -- this is done either to find a simple forbidden structure or to check a possible near cleaner for an amenable odd hole. Therefore optimizations of the path enumerating algorithm were also helpful in speeding up the CCLSV implementation.

@@

Another algorithm for which we found major speedups is the algorithm to generate all possible near-cleaners (\Cref{alg:listNearCleaners}). We optimized it by using a a \texttt{dynamic\_bitset} from the boost library \cite{boost}. It is a data structure to represent a set of bits, that also allows for fast bitwise operators that one can apply to builtin integers. At the very end of \Cref{alg:listNearCleaners} a set $\mathcal{R}$ is constructed, which is a set of all possible unions of pairs $X_i$ and $N_j$. We used \texttt{dynamic\_bitset} to represent each of the sets $X_i$ and $N_j$, and the elements of the set $\mathcal{R}$. This significantly speeds up the calculation of $\mathcal{R}$ (line \ref{line:listNCcalcR}). With this modification, the speedup of \Cref{alg:listNearCleaners} was about 20\%.

After these optimizations, checking each potential near-cleaner by \Cref{alg:testNearCleaner} is by far the biggest bottleneck of the CCLSV algorithm. We didn't find any major optimizations there, but the \Cref{alg:testNearCleaner} has a good potential for parallelization, which we explore in \Cref{sec:CUDA}.

@@

CUDA is a parallel computing platform model created by Nvidia and executed on Nvidia GPUs. It allows to run a program  by many CUDA threads at once. Although usually described by the the PRAM models, implementing PRAM algorithms in CUDA architecture is not as straightforward as implementing RAM algorithms on a CPU. Some of the causes are: synchronization between threads is very costly, there are multiple memory models to choose from, threads run divided in blocks and it is possible to efficiently share memory within blocks and copying memory from and back to CPU suffers from a large latency. What is more, there are no standard dynamic structures in the PRAM model, let alone CUDA, that are taken for granted in the RAM model, such as hashing maps or sets. 
% We used a library called \emph{moderngpu} for our implementation, because it allows a slightly cleaner code.

@@

We posed a question if it would be profitable to use GPU for the problem of recognizing Berge graphs. On one hand the graphs we are working on are small and our time complexity so large that a speedup from massively parallel architecture could be profitable. On the other hand, the algorithm is complex and not easy to parallelize. Therefore we decided to analyze the  CCLSV algorithm and parallelize parts that would benefit from it the most.

@@

Let us recall the \Cref{alg:testNearCleaner}. It calculates array $R$ of shortest paths, then for each 3-vertex path and an additional vertex it does $O(1)$ checks to see if they along with two paths from $R$ give us an odd hole. We will parallelize the work after calculating $R$, that is, lines \ref{line:cudaStart}-\ref{line:cudaEnd}. Let us notice, that for all $X$s, all 3-vertex paths enumerated in line \ref{line:3vertex} are the same. We calculate them beforehand, then each CUDA thread receives one 3-vertex path $x_1$-$x_2$-$x_3$ and an additional vertex $y_1$ and performs the required checks in lines \ref{line:cudaWorkStard}-\ref{line:cudaWorkEnd}. This is almost perfect scenario for the GPU -- we do a simple SIMD\footnote{Single Instruction Multiple Data, meaning performing same operations on multiple data points in parallel} work in parallel, without a lot scattered memory access.

The part of the CCLSV implemented on the GPU achieved speedup of up to 15x. This gave us a speedup of up to 9x for overall CCLSV algorithm, depending on a class of graph we benchmark on.
